---
title: "Analysis of Disparities"
output: html_notebook
---
Read in data

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(tidyr)
library(epitools)
library(lme4)
options(scipen=10000)
```

```{r}
all_data <- read_csv("Data/data_May3.csv")
```

Group City, and create quintiles of each variable
```{r}
all_data_quint <- all_data %>% ungroup() %>%
  group_by(city_name) %>% 
  select(city, zip:pct_unemp, pct_diabetes, pct_hbp, total_pop, tests, positives, deaths) %>%
  mutate_at(vars(mhi:pct_hbp), ~factor(findInterval(., c(-Inf,
                               quantile(., probs=c(0.2, 0.4, 0.6, 0.8), na.rm = TRUE), Inf)),
                               labels=c("Q1","Q2","Q3","Q4", "Q5"))) %>%
  pivot_longer(mhi:pct_hbp, names_to="var", values_to="value")


```

Calculate rates and CIs
```{r}
quint_summaries <- all_data_quint %>% ungroup() %>%
  group_by(city, var, value) %>%
  summarize(total_pop=sum(total_pop, na.rm = TRUE),
            tests = sum(tests, na.rm=TRUE),
            positives = sum(positives, na.rm=TRUE),
            deaths = sum(deaths, na.rm=TRUE)) %>%
  mutate(tests=ifelse(city=="Detroit" | city == "Seattle", NA, tests),
         deaths = ifelse(!(city=="Seattle" | city == "Chicago"), NA, tests)
  ) %>%
  mutate(pos_rate = positives/tests,
         test_rate = tests/total_pop,
         case_rate = (positives/total_pop)*100000,
         death_rate = (deaths/total_pop)*100000) %>%
  
  mutate(pos_rate_lci = ifelse(!is.na(pos_rate), 
                               pois.exact(positives, tests)[[4]],
                               NA),
         pos_rate_uci = ifelse(!is.na(pos_rate), 
                               pois.exact(positives, tests)[[5]],
                               NA),
         test_rate_lci = ifelse(!is.na(test_rate), 
                               pois.exact(tests, total_pop)[[4]],
                               NA),
         test_rate_uci = ifelse(!is.na(test_rate), 
                               pois.exact(tests, total_pop)[[5]],
                               NA),
         case_rate_lci = ifelse(!is.na(case_rate), 
                               (pois.exact(positives, total_pop)[[4]]*100000),
                               NA),
         case_rate_uci = ifelse(!is.na(case_rate), 
                               (pois.exact(positives, total_pop)[[5]]*100000),
                               NA),
         death_rate_lci = ifelse(!is.na(death_rate), 
                               (pois.exact(deaths, total_pop)[[4]]*100000),
                               NA),
         death_rate_uci = ifelse(!is.na(death_rate), 
                               (pois.exact(deaths, total_pop)[[5]]*100000),
                               NA))


```


Plot of all Q1 Q5 comparison of all variables in all cities
```{r}
ggplot(filter(quint_summaries, value=="Q1" | value=="Q5"), aes(group=city)) +
  geom_point(aes(x=value, y=case_rate, color=city), size=1) +
  geom_line(aes(x=value, y=case_rate, color=city, group = city), alpha=0.4, size=0.8) +
  geom_errorbar(aes(x = value, ymin=case_rate_lci, ymax=case_rate_uci, color=city), alpha=0.2) +
  scale_color_manual(values=c("#FF0000", "#00A08A", "#F2AD00", "#F98400", 
                                "#5BBCD6", "#046C9A")) +
  facet_wrap(vars(var), labeller = labeller(var = 
    c("mhi" = "Median HH Income",
      "pct_col" = "% non-white",
      "pct_diabetes" = "Diabetes prev",
      "pct_essworkers" = "% essential workers",
      "pct_hbp" = "High BP Prev",
      "pct_new_imm" = "% Entered in since 1990",
      "pct_non_cit" = "% Non citizens",
      "pct_overcrowded" = "% overcrowded",
      "pct_pov" = "% in poverty",
      "pct_old" = "% over 60 yrs",
      "pct_transit" = "% taking transit to work",
      "pct_unemp" = "% unemployed"))) + xlab("Quintile") + ylab("Cases per 100,000") +
  theme_minimal() 
```

Create long format of all data

```{r}
all_data_long <- all_data %>% 
  select(city, zip:pct_unemp, pct_diabetes, pct_hbp, total_pop, tests, positives, deaths) %>%
  pivot_longer(mhi:pct_hbp, names_to="var", values_to="value") %>%
  mutate(case_rate = (positives/total_pop)*100000) %>%
  mutate(death_rate = (deaths/total_pop)*100000) %>%
  mutate(pos_rate = (positives/tests))


```

Plot linear trends for all data 
```{r}
ggplot(filter(all_data_long), aes(group=city)) +
  geom_point(aes(x=value, y=case_rate, color=city), alpha=0.2, size=0.5) +
   stat_smooth(aes(x=value, y=case_rate, color=city, group=city),
              method="lm", se = FALSE) +
  scale_color_manual(values=c("#FF0000", "#00A08A", "#F2AD00", "#F98400", 
                                "#5BBCD6", "#046C9A")) +
  facet_wrap(vars(var), scales = "free_x", labeller = labeller(var = 
    c("mhi" = "Median HH Income",
      "pct_col" = "% non-white",
      "pct_diabetes" = "Diabetes prev",
      "pct_essworkers" = "% essential workers",
      "pct_hbp" = "High BP Prev",
      "pct_new_imm" = "% Entered since 1990",
      "pct_non_cit" = "% Non citizens",
      "pct_overcrowded" = "% overcrowded",
      "pct_pov" = "% in poverty",
      "pct_old" = "% over 60 yrs",
      "pct_transit" = "% taking transit to work",
      "pct_unemp" = "% unemployed"))) + xlab("Variable") + ylab("Cases per 100,000") +
  theme_minimal() 
```

Create dataset to fit multilevel models
```{r}
model_data <- all_data %>% ungroup() %>%
  group_by(city_name) %>% 
  select(city, zip:pct_unemp, pct_diabetes, pct_hbp, total_pop, tests, positives, deaths) %>%
  mutate_at(vars(mhi:pct_hbp), ~factor(findInterval(., c(-Inf,
                               quantile(., probs=c(0.2, 0.4, 0.6, 0.8), na.rm = TRUE), Inf)),
                               labels=c("Q1","Q2","Q3","Q4", "Q5")))
```


Loop through variables and fit individual models for each variable and store Q5/Q1 RR and signficance

```{r}
#intialize empty dataframe
results <- data.frame(var=NA, Q1=NA, Q5=NA, RR=NA, Sig=NA, Pval=NA)

for (i in names(model_data[,4:15])){
  #initialize empty dataframe
  results1 <- data.frame(var=NA, Q1=NA, Q5=NA, RR=NA, Sig=NA, Pval=NA)
  
  #Formula with random intercepts for each city and poisson model
  formula = as.formula(paste("positives ~ 1 +", i, 
                             "+ (1|city) + offset(log(total_pop))"))
  model = glmer(formula, family="poisson", data=model_data)
  
  results1$var <- i
  results1$Q1 = exp(data.frame(coef(summary(model)))[1,]$Estimate)
  results1$RR = exp(data.frame(coef(summary(model)))[5,]$Estimate)
  results1$Q5 = results1$Q1 * results1$RR
  results1$Sig = data.frame(coef(summary(model)))[5,]$Pr...z.. < 0.05
  results1$Pval = round(data.frame(coef(summary(model)))[5,]$Pr...z..,4)
  results <- rbind(results, results1)
}
```

```{r}
results
```


```{r}
model_all = glmer(positives ~ 1 + pct_col + pct_diabetes + pct_essworkers + pct_hbp + pct_new_imm + pct_overcrowded + pct_pov + pct_old + pct_transit + pct_unins + (1|city) + offset(log(total_pop)), family="poisson", data=model_data)
```

```{r}
summary(model_all)
```

```{r}
data_export <- all_data %>% ungroup() %>%
  group_by(city_name) %>% 
  select(city, zip:pct_non_cit, pct_diabetes, pct_hbp, total_pop, tests, positives, deaths)
```
Writing file with data
```{r}
write_csv(data_export, "~/COVID_19_Research/US_cities_data.csv")
```

