---
title: "Merging Case, ACS and 500 Cities Data"
output: html_notebook
---

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(tidycensus)
library(tidyr)
options(scipen=10000)
```


Get case data by zip, and zip lookup table
```{r}

city_zips <- read_csv("Data/zip_lookup.csv")
cases_zip_data <- read_csv("Data/All_cities_zip.csv") %>% select(1:5) 
```

Get ACS estimates for each ZCTA involved

% crowding, 
% Population of Color
% Population below poverty
% Essential workers
% elderly
% lacking health insurance
% public transit
% unemployed

Get list of variables from ACS 2014-18 at ZCTA level.
```{r}
vars <-load_variables(year=2018, "acs5")
```

Know what I want
Get and set up a Census API from https://api.census.gov/data/key_signup.html

```{r}
census_api_key("c03dbff727c9699cb8ccb3282d07867739a8afef")


vars <- c(# Essential industry workers
  "C24010_001", "C24010_012", "C24010_016", "C24010_020", "C24010_021", "C24010_024",
          "C24010_025",
          "C24010_030", "C24010_034", 
          "C24010_048", "C24010_052", "C24010_056", "C24010_057", "C24010_060", "C24010_061",
          "C24010_066", "C24010_070", 
          # median household income
          "B19013_001",
          # below poverty, and # total
          "B17001_002", "B17001_001",
          # overcrowding denominator
          "B25014_001", 
          #overcrowding 1 to 1.5
          "B25014_005", "B25014_011",
          #overcrowding 1.5 to 2
          "B25014_006", "B25014_012",
          #overcrowding 2more
          "B25014_007", "B25014_013",
          # total population, # total non white non hispanic, 
          "B01003_001", "B01001H_001",
          #  60-70 male
          "B01001_018", "B01001_019", "B01001_020", "B01001_021", 
          # > 70 male
          "B01001_022", "B01001_023", "B01001_024", "B01001_025",
          # 60-70 female
          "B01001_042", "B01001_043", "B01001_044", "B01001_045", 
          # > 70 female
          "B01001_046", "B01001_047", "B01001_048", "B01001_049", 
          # public transit excluding taxicab
          "B08006_001","B08006_008",
          # health insurance (uninsured)
          #"B27010_001","B27010_017", "B27010_033", "B27010_050", "B27010_066",
          #Total, #Entered after 1990, foreign born
          "B05005_001", "B05005_004", "B05005_009", "B05005_014",
          #Total, #Unemployed,
          "C18120_001", "C18120_006",
          #Total, #Not citizen
          "B05001_001", "B05001_006"
          ) 



```

Get the data
```{r}
covar_zip_data_raw <- get_acs(geography = "zcta",
                      variables=vars, cache_table = TRUE) 


```


Get population data at census tract level as well to be able to aggregate 500 Cities data up to ZCTA level.
```{r}
tract_pop_data <- get_acs(geography = "tract", variables = "B01003_001",
                          state=c("Washington", "Illinois", "Michigan", "New York", "Pennsylvania", "Massachusetts"),
                          year=2018) 

tract_pop_data <- tract_pop_data %>%
  select(GEOID, variable, estimate) %>% 
  spread(variable, estimate) %>%
  rename(total_pop=B01003_001) %>%
  select(GEOID, total_pop) %>% mutate(GEOID = as.character(GEOID))

```


Transform the ACS data to create variables of interest
```{r}
covar_zip_data <- covar_zip_data_raw %>%
  select(GEOID, variable, estimate) %>% 
  filter(GEOID %in% city_zips$ZCTA5) %>%
  spread(variable, estimate) %>% 
  rowwise() %>% 
  
  # create indicators
  mutate(mhi=B19013_001,
         
         pct_pov = B17001_002/B17001_001,
         
         pct_col =  (B01003_001 - B01001H_001) / B01003_001,
         
         total_pop = B01003_001,
         
         #pct_unins=sum(c(B27010_017, B27010_033, B27010_050, B27010_066))/B27010_001,
         
         pct_essworkers=sum(c(C24010_012, C24010_016, C24010_020, C24010_021, C24010_024, C24010_025,
          C24010_030, C24010_034, 
          C24010_048, C24010_052, C24010_056, C24010_057, C24010_060, C24010_061,
          C24010_066, C24010_070))/C24010_001,
         
         pct_overcrowded=sum(c(B25014_007, B25014_006,B25014_005,
                                B25014_011,B25014_012, B25014_013))/B25014_001,
         
         pct_transit=B08006_008/B08006_001,
         
         pct_old=sum(c(B01001_018, B01001_019, B01001_020, B01001_021, 
          B01001_022, B01001_023, B01001_024, B01001_025,
          B01001_042, B01001_043, B01001_044, B01001_045,
          B01001_046, B01001_047, B01001_048, B01001_049)/B01003_001),
         
         pct_new_imm=(B05005_004+B05005_009+B05005_014)/B05005_001,
         pct_non_cit=B05001_006/B05001_001,
         
         pct_unemp= C18120_006/C18120_001,
         
         GEOID=as.character(GEOID)) %>% 
  select(GEOID, mhi, pct_pov, pct_col, total_pop, pct_essworkers, pct_overcrowded, pct_transit,
         pct_old, pct_new_imm, pct_non_cit, pct_unemp)
```

Get health data downloaded from RWJF City Health Dashboard website (500 Cities). 
https://www.cityhealthdashboard.com/
I dowloaded the state files from the RWJF website, but they also have an API
Used prevalence of high blood pressure and diabetes from their data

The 500 Cities data is at census tract level, so need the ZCTA - CT relationship file from the census website: https://www.census.gov/geographies/reference-files/2010/geo/relationship-files.html#par_textimage_674173622  (zcta_tract_rel_10.csv)

The population weighted mean is taken to convert the CT-level data to ZCTA-level data
```{r}

geog_file <- read_csv("Data/zcta_tract_rel_10.csv") %>% select(ZCTA5, GEOID) %>% mutate(GEOID=as.character(GEOID))
health_data_raw <- do.call(rbind,
          lapply(paste0("Data/Health_data/", (list.files(path = "Data/Health_data/"))), read.csv)) %>%
  filter(city_name %in% c("New York", "Chicago", "Seattle", "Philadelphia",  "Detroit", "Boston")) %>%
  filter(metric_number %in% c(6,9)) %>%
  mutate(stcotr_fips = as.character(stcotr_fips)) %>%
  left_join(geog_file, by=c("stcotr_fips"="GEOID")) %>%
  left_join(tract_pop_data, by=c("stcotr_fips"="GEOID")) %>%
  right_join(city_zips, by=c("ZCTA5")) %>%
  group_by(city_name, metric_name, zip)
  
  
health_data <- health_data_raw %>%
  summarise(est_weight = weighted.mean(est, total_pop, na.rm = TRUE),
            pop_total=sum(total_pop)) %>%
  mutate(est_weight=ifelse(is.nan(est_weight), 0, est_weight)) %>%
  spread(metric_name, est_weight) %>% 
  rename(pct_diabetes = Diabetes, pct_hbp = `High blood pressure`) %>%
  select(zip, city_name, pct_diabetes, pct_hbp)
  
```

Merge data. For Boston, where cases are reported for some zipcodes in an aggregated way, the population weighted mean of the variables are taken.
```{r}
all_data <- covar_zip_data %>% 
  left_join(city_zips, by=c("GEOID"="ZCTA5")) %>% filter(GEOID != "98195") %>%
  group_by(zip) %>%
  summarise(mhi = weighted.mean(mhi, total_pop, na.rm = TRUE),
            pct_pov = weighted.mean(pct_pov, total_pop, na.rm = TRUE),
            pct_col = weighted.mean(pct_col, total_pop, na.rm = TRUE),
            #pct_unins = weighted.mean(pct_unins, total_pop, na.rm = TRUE),
            pct_essworkers = weighted.mean(pct_essworkers, total_pop, na.rm = TRUE),
            pct_overcrowded = weighted.mean(pct_overcrowded, total_pop, na.rm = TRUE),
            pct_transit = weighted.mean(pct_transit, total_pop, na.rm = TRUE),
            pct_old = weighted.mean(pct_old, total_pop, na.rm = TRUE),
            pct_new_imm = weighted.mean(pct_new_imm, total_pop, na.rm = TRUE),
            pct_non_cit = weighted.mean(pct_non_cit, total_pop, na.rm = TRUE),
            pct_unemp = weighted.mean(pct_unemp, total_pop, na.rm = TRUE),
            total_pop = sum(unique(total_pop), na.rm = TRUE)) %>%
  left_join(health_data, by=c("zip")) %>% 
  filter(!is.na(city_name)) %>%
  left_join(cases_zip_data, by=c("zip"="custom_zip"))
```


Write data

```{r}
write_csv(all_data, "Data/data_May3.csv")
```


